---
title: "Adams_mDMC"
author: "TP"
format: html
editor: visual
---

## Zhang Actual Evapotranspiration

This script will utilize data from ClimateBC to calculate Zhang's AET (Zhang, 2004).

## Loading necessary variables

Zhang's equation utilizes PET and Precipitation data. Hargreaves PET and Precip data for the Adams Wtershed is accessible through the ClimateBC v7.4.

```{r}
library(terra)

folder_path <- ("C:/Users/taypo/OneDrive/Documents/Grad Studies/Graduate Research/mDMC/Adams Raster Files/Monthly Climate Rasters")

year_folders <- dir(folder_path, full.names = TRUE)

monthly_dirs <- lapply(year_folders, list.files, full.names = TRUE,
                       pattern = "^Eref[0-9]{2}\\.asc$|^PPT[0-9]{2}\\.asc$")

iteration_table <- lapply(1:12, seq, to = 24, by = 12)

sorted_data_lists <- lapply(monthly_dirs, function(x) {
                            setNames(lapply(iteration_table, function(y) {
                                            rast(x[y])
                            }), month.name)
})

sorted_data_lists <- setNames(sorted_data_lists, basename(year_folders))

```

## Setting up Actual Evapotranspiration calculation

Now the PET (Eref) and Precip (PPT) data are loaded into sorted lists, let's set up the calculation for AET. The w coefficient of 2.84 was chosen based on the optimized w for a forested catchment (Zhang 2004).

```{r}
calculate_aet <- function(ppt, eref) {
  aet <- ppt * (1 + 2.84 * (eref/ppt)) / (1 + 2.84 * (eref/ppt) + ppt/eref)
  return(aet)
}
```

## Calculating AET for each year

Looping through each year and each month, extracting the necessary SpatRasters and calculating AET.

```{r}
# Loop through each year
for (year in names(sorted_data_lists)) {
  # Extract year from the year folder name
  year_num <- as.numeric(sub("M", "", strsplit(year, "_")[[1]][2]))
  
  # Loop through each month
  for (month in names(sorted_data_lists[[year]])) {
    # Construct the layer names based on the month
    ppt_name <- paste0("PPT", sprintf("%02d", match(month, month.name)))
    eref_name <- paste0("Eref", sprintf("%02d", match(month, month.name)))
    
    # Extract raster layers for the current month
    ppt <- sorted_data_lists[[year]][[month]][[ppt_name]]
    eref <- sorted_data_lists[[year]][[month]][[eref_name]]

    # Calculate AET
    aet <- calculate_aet(ppt, eref)
    
    # Define the filename for saving
    month_num <- match(month, month.name) # Get numeric month
    filename <- paste0("AET_", year_num, "_", sprintf("%02d", month_num), ".tif")
    
    # Save the raster
    terra::writeRaster(aet, filename, overwrite = TRUE)
  }
}
```

## Extracting AET values

We need to extract values from each of the monthly raster layers containing AET values. AET values are in mm, so will need to sum values from each cell for each month, add them to a single dataframe, and then export to a xlsx file.

```{r}
library(lubridate)  # For date operations
library(openxlsx)

file_path <- "C:/Users/taypo/OneDrive/Documents/GitHub/Adams_glacier_study/Zhang_AET"

aet_files <- list.files(path = file_path, full.names = TRUE)

monthly_aet <- data.frame(month = character(), aet = numeric())

# Loop through each raster file
for (aet_file in aet_files) {
  # Extract year and month from the raster file name
  filename_parts <- strsplit(basename(aet_file), "_")[[1]]
  year <- as.numeric(filename_parts[2])
  month <- as.numeric(sub("\\.tif$", "", filename_parts[3]))
  
  # Remove ".tif" extension from month
  file_date_str <- paste(year, month, "01", sep = "-")
  
  # Create a date object
  file_date <- as.Date(file_date_str)
  
  # Read the raster
  rast <- rast(aet_file)
  
  # Extract values and sum them
  ave_value <- mean(values(rast), na.rm = TRUE)
  
  # Add the summed values to the data frame
  monthly_aet <- rbind(monthly_aet, data.frame(month = file_date, aet = ave_value))
}

# Write the data frame to an Excel file
write.xlsx(monthly_aet, "monthly_aet.xlsx", rowNames = FALSE)

```

## Calculating Effective Precipitation

Effective precipitation is actual precipitation less actual evapotranspiration. Need to load all PPT monthly raster files, convert them to a df and combine the ppt and aet dataframes to make the calculation.

```{r}
library(tidyverse)

folder_path <- ("C:/Users/taypo/OneDrive/Documents/Grad Studies/Graduate Research/mDMC/Adams Raster Files/Monthly Climate Rasters")

year_folders <- dir(folder_path, full.names = TRUE)

ppt_dirs <- lapply(year_folders, list.files, full.names = TRUE,
                       pattern = "^PPT[0-9]{2}\\.asc$")

monthly_ppt <- data.frame(month = character(), precip = numeric())

iteration_ppt <- lapply(1:12, seq, to = 12, by = 12)

sorted_ppt_list <- lapply(ppt_dirs, function(x) {
                            setNames(lapply(iteration_ppt, function(y) {
                                            rast(x[y])
                            }), month.name)
})

sorted_ppt_list <- setNames(sorted_ppt_list, basename(year_folders))

for (year in names(sorted_ppt_list)) {
  for (month in names(sorted_ppt_list[[year]])) {
    ppt_name <- paste0("PPT", sprintf("%02d", match(month, month.name)))
    rast_ppt <- sorted_ppt_list[[year]][[month]][[ppt_name]]
    ave_ppt <- mean(values(rast_ppt), na.rm = TRUE)  
    monthly_ppt <- rbind(monthly_ppt, data.frame(month, precip = ave_ppt))
  }
}

write.xlsx(monthly_ppt, "monthly_ppt.xlsx", rowNames = FALSE)

```

## Combing data

Now we need to add precip data, aet, and discharge data into one dataframe. The discharge data is recorded in daily increments, so these values will need to be summed by month. Furthermore, the discharge is in m\^3/s; therefore, this will need to be converted into mm.

```{r}
library(readxl)
library(dplyr)
library(tidyverse)
library(ggplot2)

adams_discharge <- read_excel("C:/Users/taypo/OneDrive/Documents/Grad Studies/Graduate Research/mDMC/Excel data/Adams_Discharge_1970_2021.xlsx")

# Convert date column to Date format
adams_discharge$Date <- as.Date(adams_discharge$Date)

adams_discharge$`Discharge (mm/day)` <- ((adams_discharge$`Discharge (m3/s)` / 3137132572.000) * 86400 * 1000)

# Now, you can proceed with the mutate operation
adams_discharge <- adams_discharge %>%
   mutate(year = lubridate::year(Date),
          month = lubridate::month(Date, label = TRUE))

# Group by year and month, then summarize the values
monthly_discharge <- adams_discharge %>%
  group_by(year, month) %>%
  summarise(discharge = sum(`Discharge (mm/day)`, na.rm = TRUE))

# Output the result
print(monthly_discharge)

monthly_aet <- read_excel("C:/Users/taypo/OneDrive/Documents/GitHub/Adams_glacier_study/monthly_aet.xlsx")
monthly_ppt <- read_excel("C:/Users/taypo/OneDrive/Documents/GitHub/Adams_glacier_study/monthly_ppt.xlsx")

effective_precip <- cbind(monthly_aet, monthly_ppt)
effective_precip <- effective_precip[,-3]

effective_precip <- effective_precip %>% 
  mutate(ppt_effective = precip - `aet`)

effective_precip_70_21 <- head(effective_precip, -12)

mDMC_monthly <- cbind(monthly_discharge, effective_precip_70_21)
mDMC_monthly <- mDMC_monthly[,-2]
names(mDMC_monthly)[3] <- "month"
mDMC_monthly <- mDMC_monthly[, c(1, 3, 4, 5, 6, 2)]

mDMC_monthly$month <- as.Date(mDMC_monthly$month, format = "%Y-%m-%d")

# Define the start and end dates of each hydrological year
hydro_year_start <- as.Date(paste0(year(mDMC_monthly$month), "-10-01"))
hydro_year_end <- as.Date(paste0(year(mDMC_monthly$month) + 1, "-09-30"))

# Create a new variable indicating the hydrological year
mDMC_monthly$hydro_year <- ifelse(mDMC_monthly$month >= hydro_year_start & mDMC_monthly$month <= hydro_year_end,
                        year(mDMC_monthly$month),
                        year(mDMC_monthly$month) - 1)

mDMC_hydro_yr <- mDMC_monthly[,- c(1,2)]

mDMC_hydro_yr <- mDMC_hydro_yr %>%
  group_by(hydro_year) %>%
  summarise(across(everything(), sum))

# Remove hydroyears with missing values. In this case, hydro year 1969, 1995, and 1996.
mDMC_hydro_yr <- mDMC_hydro_yr[-c(1,27,28),]

mDMC_hydro_yr$cumulative_precip <- cumsum(mDMC_hydro_yr$ppt_effective)
mDMC_hydro_yr$cumulative_discharge <- cumsum(mDMC_hydro_yr$discharge)

write.xlsx(mDMC_hydro_yr, "mDMC_hydro_yr.xlsx", rowNames = FALSE)

ggplot(mDMC_hydro_yr, aes(x = cumulative_precip, y = cumulative_discharge)) +
  geom_point() +
  labs(x = "Cumulative Precipitation", y = "Cumulative Discharge", title = "Adams mDMC")

```

```{r}
library(tidyverse)      
library(changepoint)    
library(zoo)
library(segmented)

mDMC_lm <- lm(cumulative_discharge ~ cumulative_precip, data = mDMC_hydro_yr)
summary(mDMC_lm)

davies.test(mDMC_lm, k=10)

mDMC_hydro_yr$pettit <- mDMC_hydro_yr$discharge - mDMC_hydro_yr$ppt_effective

pettit_test <- pettitt.test(mDMC_hydro_yr$pettit)
summary(pettit_test)

```
