---
title: "Adams hydrostn area & elevation"
format: html
editor: visual
---

## Defining study area

### Hydrological Boundary

First we must define the area of interest. The Adams hydrometric station near Squilax (08LD001) is at the outflow of Adams Lake; therefore, we will need to spatially define the watershed to encompass all land from which water flows to this point.

```{r}
install.packages("bcdata")

library(bcdata)
library(sf)

hydro_area_search <- bcdc_search("hydrometric")
View(hydro_area_search)

bc_hydrostns <- bcdc_query_geodata(hydro_area_search[[3]]$id, crs = 3005) %>% 
  collect()

adams_hydrostn <- bc_hydrostns %>% 
  filter(SOURCE_NAME == "ADAMS RIVER")

st_write(adams_hydrostn, "adams_hydrostn.shp")
```

### Watershed Digital Elevation Model

```{r}
install.packages("bcmaps")
install.packages("mapview")

library(bcmaps)
library(mapview)
library(terra)

adams_hydrostn_dem <- cded_terra(adams_hydrostn)

crs(adams_hydrostn_dem)

adams_dem_albers <- project(adams_hydrostn_dem, "epsg:3005")
res(adams_dem_albers)

resamp_grid <- rast(ext(adams_dem_albers), res = 20, crs = "epsg:3005")
adams_dem_albers <- resample(adams_dem_albers, resamp_grid)
plot(adams_dem_albers)

adams_dem_mask <- mask(adams_dem_albers, vect(adams_hydrostn))

writeRaster(adams_dem_mask, "Adams_hydrostn_DEM.tif", overwrite = TRUE)
```

### Daytime Length

First, we must calculate daylight hours for the Adams Watershed. As the calculation is only refined to the nearest minute within the DMS coordinate system, we will use

```{r}

```
